Refactoring `nls()`: a constructive critique

### 2023-01-28 RefactoringNLS/RefactoringNLS.Rmd
note on profiling
Incorporating AB edits

---
title: "Refactoring the `nls()` function in R"
author: 
   - John C. Nash \thanks{ retired professor, Telfer School of Management, University of Ottawa}
   - Arkajyoti Bhattacharjee \thanks{Department of Mathematics and Statistics, Indian Institute of Technology, Kanpur}
date: "2022-6-15"
output: 
    pdf_document:
        keep_tex: true
        toc: true
bibliography: ../BibSupport/ImproveNLS.bib
link-citations: yes
linkcolor: red
urlcolor: blue
---

@@

Important considerations are the usability and  maintainability of
the code base that provides the functionality `nls()` claims to offer.
Our work suggests that the
existing code makes maintenance and improvement very difficult,
with legacy applications and examples blocking some important updates.
Discussion of these matters is relevant to improving R generally, as
well as its nonlinear estimation tools.

@@

- handles formulas that include R functions, even ones which call calculations in
  other programming languages
- allows data to be weighted or subset
- can estimate bound constrained parameters
- provides a mechanism for handling partially linear models
- permits parameters to be indexed over a set of related data
- produces measures of variability (i.e., standard error estimates) for the estimated parameters
- has related profiling capabilities for exploring the likelihood surface as parameters are changed
- links to a number of pre-coded ("selfStart") models

@@

Adding a small quantity to the loss function before dividing avoids trouble. 
In 2021, one of us (J. Nash) proposed that `nls.control()` have an additional parameter `scaleOffset` 
with a default value of zero. Setting it to a small number -- 1.0 is 
a reasonable choice --
allows small-residual problems (i.e., near-exact fits) to be dealt with easily. We call this the
**safeguarded relative offset convergence criterion**. The default value gives the legacy behaviour.
We are pleased to note that this improvement is now in the R distributed code as of version 4.1.0.

@@

The Jacobian used by `nlsr::nlxb()` is, by default, computed from analytic expressions, 
while that
of `nls()` mostly uses a numerical approximation. Some selfStart models do provide
"gradient" information. `minpack.lm::nlsLM()` invokes `numericDeriv()` in its local 
version of `nlsModel()`, but it appears to use an internal approximate Jacobian code from 
the original Fortran `minpack` code, namely, `lmdif.f`. Such differences in approach
can lead to different behaviour, though in our experience differences are usually minor.

@@

Note that the selfStart models in the base R `./src/library/stats/R/zzModels.R` file 
and in package `nlraa` (@MiguezNLRAA2021) may
provide the Jacobian in the "gradient" attribute of the "one-sided" formula that defines
each model, and these Jacobians may be the analytic forms. The `nls()` function, after
computing the "right hand side" or `rhs` of the residual, then checks to see if the
"gradient" attribute is defined, and, if not, uses `numericDeriv` to compute a Jacobian
into that attribute. This code is within the `nlsModel()` or `nlsModel.plinear()`
functions. The use of analytic Jacobians
almost certainly contributes to the good performance of `nls()` on selfStart models.

@@

Our concern with `na.action` is that users may be unaware of the effects of an option
setting they may not even be aware has been set. Is `na.fail` a safer default?

@@

Nonlinear estimation algorithms are almost all iterative and need a set of starting
parameters. `nls()` offers a special class of modeling functions called **selfStart** 
models. There are a number of these in base R (`./src/library/stats/R/zzModels.R`)
and others in R packages such as CRAN package `nlraa` (@MiguezNLRAA2021), as well as 
the now-archived package `NRAIA`. Unfortunately, some **selfStart** codes
entangle the calculation of starting values for parameters
with the particulars of the `nls()` code. Though there is a `getInitial()` function,
this is not easy to use to simply compute the initial parameter estimates outside of
`nls()`, in part 
because it may call that function. Such circular references are, in our view, dangerous.
Moreover, we believe that it would be helpful to have selfStart models that allow
users to explicitly provide values other than those suggested for the starting parameters.

Other concerns are revealed by the example below. Here, the `SSlogis` selfStart function 
is used to generate a set
of initial parameters for a 3-parameter logistic curve. The form used by `SSlogis`
is $y \,\sim\, Asym\,/\, (1 \,+\, exp((xmid\,-\,tt)\,/\,scal))$, but we show how the starting parameters
for this model can be transformed to those of 
another form of the model, namely, $y \,\sim\, b1\,/\,(1 \,+\, b2\,*\,exp(-b3\,*\,t))$.

@@

In the event that a selfStart model is not available, `nls()` sets all the starting parameters 
to 1. This is, in our view, tolerable, but could be improved by using a set of values
that are all slightly different, which, in the case of the example model $y \,\sim\, a \,*\, exp(-b \,*\, x) + c\,*\,exp(-d \,*\, x)$
would avoid a singular Jacobian because $b$ and $d$ were equal in value. 
A sequence like  1.0, 1.1, 1.2, 1.3 for the four parameters 
could be provided quite simply instead of all 1's.

@@

### Strategic issues in selfStart models

By providing starting values that are likely to be reasonable and by including 
Jacobian code in the selfStart model code, some of the deficiencies of the 
Gauss-Newton algorithm are mitigated. However, creating such functions is a lot 
of work, and their
documentation (file `./src/library/stats/man/selfStart.Rd`) is quite complicated.
We believe that the focus should be placed on getting good initial parameters,
that is `getInitial()` function, though avoiding the current calls back to `nls()`.
Interactive tools, such as "visual fitting" (@nash1996nonlinear) might be
worth considering.

@@

Let us compare the `nls()` result with that from 
`nlsr::nlxb()`, with ostensibly solves the same problem:

@@

```
 $ lower       : num [1:3] -Inf -Inf -Inf
 $ upper       : num [1:3] Inf Inf Inf
 $ maskidx     : int(0) 
 $ weights     : NULL
 $ formula     :Class 'formula'  language y ~ Asym/(1 + exp((xmid - tt)/scal))
  .. ..- attr(*, ".Environment")=<environment: R_GlobalEnv> 
 - attr(*, "class")= chr "nlsr"
```
The result object from `nlsr::nlxb()` is still much smaller than the one `nls()` returns. 
Moreover, `nlxb`
explicitly returns the sum of squares as well as the residual vector and Jacobian. 
The counts of evaluations are also returned.
(Note that the singular values of the Jacobian are
actually computed in the `print` and `summary` methods for the result.) 
As we made our comparisons, We noted several potential updates to the `nlsr` documentation
as well as that for `nls()`.

@@

The variable projection method (@Golub1973, @OlearyRust13) is generally much more 
effective then general approaches in finding good solutions to nonlinear least squares 
problems when some of the parameters appear linearly. However, setting up the 
calculations, that is, identifying which parameters are linear, is not a trivial
task.

@@

The nonlinear modeling model specifications are, of course, a development of linear ones.
Unfortunately, the explicit model `y ~ a * x + b` does not work with the linear modeling 
function `lm()`, which requires this model to be specified as `y ~ x`.

@@

  ## Then we use nls' indexing feature for parameters in non-linear
  ## models to use the conventional algorithm to fit a model in which
  ## alpha and beta vary with Strip.  The starting values are provided
  ## by the previously fitted model.
  ## Note that with indexed parameters, the starting values must be
  ## given in a list (with names):
  ## ?? but why use b here AND in the new formula??
  b <- coef(musc.1)
  musc.2 <- nls(Length ~ a[Strip] + b[Strip]*exp(-Conc/th), data=mm,
                start = list(a = rep(b[2], nl), b = rep(b[3], nl), th = b[1]))
  summary(musc.2)
})
```

@@

Indexed models clearly have a place in some areas of research. 
<!-- We need to be able to process models, as the example above, where the -->
<!-- model formula is  -->
We need to be able to process models such as 
`Length ~ a[Strip] + b[Strip] * exp(-Conc / th)` where `Strip` is the index. Data
`Length` and `Conc` are available for all observations and parameter `th` applies
to every fitted point, but `a[Strip]` and `b[Strip]` apply only to data indexed, which
could be an experimental run number, or some other categorization.

Given that `[ ]` are the R method to index arrays, their presence in a model
formula signals indexing. The question is how to perform the correct calculations.

@@

We anticipate that tools like `nlsr` will become more attractive as users discover
the increased robustness and general capabilities, along with the more recent documentation.
From a design point of view, a key difference in approach between `nls()` and 
`nlsr::nlxb()` is that `nls()` 
builds a large infrastructure, especially the `nlsModel()` function,  from which the Gauss-Newton
iteration can be executed and other statistical information such as profiles can be
computed, while `nlxb()` returns quite limited information, and 
computes what is needed on an "as and when" basis. This follows a path that one of us
(JN) established almost 50 years ago with the software that became @cnm79, separating
setup, solver, and post-solution analysis phases of computations. 
<!-- Here, that last  -->
<!-- phase is not part of `nlxb()`, but is the domain of other functions in the `nlsr` -->
<!-- package. -->

@@

Hans Werner Borchers has been helpful in developing the proposal for this project
and in comments on this and related work. Heather Turner co-mentored the project 
and helped guide the progress of the work. Exchanges with Fernando Miguez helped to
clarify aspects of selfStart models and improve package `nlsr` in that respect.

>>>>

---
title: "Refactoring `nls()`: a constructive critique"
author: 
   - John C. Nash \thanks{ retired professor, Telfer School of Management, University of Ottawa}
   - Arkajyoti Bhattacharjee \thanks{Department of Mathematics and Statistics, Indian Institute of Technology, Kanpur}
date: "2022-6-15"
output: 
    pdf_document:
        keep_tex: true
        toc: true
bibliography: ../BibSupport/ImproveNLS.bib
link-citations: yes
linkcolor: red
urlcolor: blue
---

@@

Important considerations are the usability and  maintainability of
the code base that provides the functionality `nls()` claims to offer.
Our work suggests that the
existing code makes maintenance and improvement very difficult,
with legacy applications and examples blocking some important updates.
Discussion of these matters is relevant to improving R generally as
well as its nonlinear estimation tools.

@@

- handles formulas that include R functions, even ones which call calculations in
  other programming languages
- allows data to be weighted or subset
- can estimate bound constrained parameters
- provides a mechanism for handling partially linear models
- permits parameters to be indexed over a set of related data
- produces measures of variability (i.e., standard error estimates) for the estimated parameters
- has related profiling capabilities for exploring the likelihood surface as parameters are changed
- links to a number of pre-coded (`selfStart`) models

@@

Adding a small quantity to the loss function before dividing avoids trouble. 
In 2021, one of us (J. Nash) proposed that `nls.control()` have an additional parameter `scaleOffset` 
with a default value of zero. Setting it to a small number -- 1.0 is 
a reasonable choice --
allows small-residual problems (i.e., near-exact fits) to be dealt with easily. We call this the
**safeguarded relative offset convergence criterion**. The default value gives the legacy behaviour.
We are pleased to note that this improvement is now in the R distributed code since version 4.1.0.

@@

The Jacobian used by `nlsr::nlxb()` is, by default, computed from analytic expressions, 
while that
of `nls()` mostly uses a numerical approximation. Some `selfStart` models do provide
"gradient" information. `minpack.lm::nlsLM()` invokes `numericDeriv()` in its local 
version of `nlsModel()`, but it appears to use an internal approximate Jacobian code from 
the original Fortran `minpack` code, namely, `lmdif.f`. Such differences in approach
can lead to different behaviour, though in our experience differences are usually minor.

@@

Note that the `selfStart` models in the base R `./src/library/stats/R/zzModels.R` file 
and in package `nlraa` (@MiguezNLRAA2021) may
provide the Jacobian in the "gradient" attribute of the "one-sided" formula that defines
each model, and these Jacobians may be the analytic forms. The `nls()` function, after
computing the "right hand side" or `rhs` of the residual, then checks to see if the
"gradient" attribute is defined, and, if not, uses `numericDeriv` to compute a Jacobian
into that attribute. This code is within the `nlsModel()` or `nlsModel.plinear()`
functions. The use of analytic Jacobians
almost certainly contributes to the good performance of `nls()` on `selfStart` models.

@@

Our concern with `na.action` is that users may be unaware of the effects of an option
setting they may not even be aware has been set. Should `na.fail` be the default?

@@

Nonlinear estimation algorithms are almost all iterative and need a set of starting
parameters. `nls()` offers a special class of modeling functions called **`selfStart`** 
models. There are a number of these in base R (`./src/library/stats/R/zzModels.R`)
and others in R packages such as CRAN package `nlraa` (@MiguezNLRAA2021), as well as 
the now-archived package `NRAIA`. Unfortunately, some **`selfStart`** codes
entangle the calculation of starting values for parameters
with the particulars of the `nls()` code. Though there is a `getInitial()` function,
this is not easy to use to simply compute the initial parameter estimates outside of
`nls()`, in part 
because it may call that function. Such circular references are, in our view, dangerous.
Moreover, we believe that it would be helpful to have `selfStart` models that allow
users to explicitly provide values other than those suggested for the starting parameters.

Other concerns are revealed by the example below. Here, the `SSlogis` `selfStart` function 
is used to generate a set
of initial parameters for a 3-parameter logistic curve. The form used by `SSlogis`
is $y \,\sim\, Asym\,/\, (1 \,+\, exp((xmid\,-\,tt)\,/\,scal))$, but we show how the starting parameters
for this model can be transformed to those of 
another form of the model, namely, $y \,\sim\, b1\,/\,(1 \,+\, b2\,*\,exp(-b3\,*\,t))$.

@@

In the event that a `selfStart` model is not available, `nls()` sets all the starting parameters 
to 1. This is, in our view, tolerable, but could be improved by using a set of values
that are all slightly different, which, in the case of the example model $y \,\sim\, a \,*\, exp(-b \,*\, x) + c\,*\,exp(-d \,*\, x)$
would avoid a singular Jacobian because $b$ and $d$ were equal in value. 
A sequence like  1.0, 1.1, 1.2, 1.3 for the four parameters 
could be provided quite simply instead of all 1's.

@@

### Strategic issues in `selfStart` models

By providing starting values that are likely to be reasonable and by including 
Jacobian code in the `selfStart` model code, some of the deficiencies of the 
Gauss-Newton algorithm are mitigated. However, creating such functions is a lot 
of work, and their
documentation (file `./src/library/stats/man/selfStart.Rd`) is quite complicated.
We believe that the focus should be placed on getting good initial parameters,
that is `getInitial()` function, though avoiding the current calls back to `nls()`.
Interactive tools, such as "visual fitting" (@nash1996nonlinear) might be
worth considering.

@@

Let us compare the `nls()` result with that from 
`nlsr::nlxb()`, which ostensibly solves the same problem:

@@

```
 $ lower       : num [1:3] -Inf -Inf -Inf
 $ upper       : num [1:3] Inf Inf Inf
 $ maskidx     : int(0) 
 $ weights     : NULL
 $ formula     :Class 'formula'  language y ~ Asym/(1 + exp((xmid - tt)/scal))
  .. ..- attr(*, ".Environment")=<environment: R_GlobalEnv> 
 - attr(*, "class")= chr "nlsr"
```
The result object from `nlsr::nlxb()` is still much smaller than the one `nls()` returns. 
Moreover, `nlxb`
explicitly returns the sum of squares as well as the residual vector and Jacobian. 
The counts of evaluations are also returned.
(Note that the singular values of the Jacobian are
actually computed in the `print` and `summary` methods for the result.) 
As we made our comparisons, we noted several potential updates to the `nlsr` documentation
as well as that for `nls()`.

@@

The variable projection method (@Golub1973, @OlearyRust13) is generally much more 
effective than general approaches in finding good solutions to nonlinear least squares 
problems when some of the parameters appear linearly. However, setting up the 
calculations, that is, identifying which parameters are linear, is not a trivial
task.

@@

The nonlinear modeling specifications are, of course, a development of linear ones.
Unfortunately, the explicit model `y ~ a * x + b` does not work with the linear modeling 
function `lm()`, which requires this model to be specified as `y ~ x`.

@@

  ## Then we use nls' indexing feature for parameters in non-linear
  ## models to use the conventional algorithm to fit a model in which
  ## alpha and beta vary with Strip.  The starting values are provided
  ## by the previously fitted model.
  ## Note that with indexed parameters, the starting values must be
  ## given in a list (with names):
  b <- coef(musc.1)
  musc.2 <- nls(Length ~ a[Strip] + b[Strip]*exp(-Conc/th), data=mm,
                start = list(a = rep(b[2], nl), b = rep(b[3], nl), th = b[1]))
  summary(musc.2)
})
```

@@

Indexed models clearly have a place in some areas of research. 
<!-- We need to be able to process models, as the example above, where the -->
<!-- model formula is  -->
We need to be able to process models such as 
`Length ~ a[Strip] + b[Strip] * exp(-Conc / th)`, where `Strip` is the index. Data
`Length` and `Conc` are available for all observations and parameter `th` applies
to every fitted point, but `a[Strip]` and `b[Strip]` apply only to data indexed, which
could be an experimental run number, or some other categorization.

Given that `[ ]` provides the R method to index arrays, the presence of square
brackets in a model
formula signals indexing. The question is how to perform the correct calculations.

@@

We anticipate that tools like `nlsr` will become more attractive as users discover
the increased robustness and general capabilities, along with the more recent documentation.
From a design point of view, a key difference in approach between `nls()` and 
`nlsr::nlxb()` is that `nls()` 
builds a large infrastructure, especially the `nlsModel()` function,  from which the Gauss-Newton
iteration can be executed and other statistical information such as profiles can be
computed, while `nlxb()` returns quite limited information, and 
computes what is needed on an "as and when" basis. This follows a path that one of us
(JN) established almost 50 years ago with the software that became @cnm79, separating
setup, solver, and post-solution analysis phases of computations. There is, however,
the wrapper `wrapnlsr()` that solves the nonlinear estimation problem with the solver
of `nlsr` then calls `nls()` to build the object of class `nls` for which profile log-likelihood
information can be computed.
<!-- Here, that last  -->
<!-- phase is not part of `nlxb()`, but is the domain of other functions in the `nlsr` -->
<!-- package. -->

@@

Hans Werner Borchers has been helpful in developing the proposal for this project
and in comments on this and related work. Heather Turner co-mentored the project 
and helped guide the progress of the work. Exchanges with Fernando Miguez helped to
clarify aspects of `selfStart` models and improve package `nlsr` in that respect.
Colin Gillespie (package `benchmarkme`) has been helpful in guiding our attempts
to succinctly summarize computing environments.
